### 2019-07-02 manuscript.Rmd
manuscript

---
author:
  - name: Nan-Hung Hsieh
    affiliation: Texas A&M University
  - name: Brad Reisfeld
    affiliation: Colorado State University 
  - name: Weihsueh A. Chiu
    affiliation: Texas A&M University
    address: >
      4458 TAMU, 
      College Station, TX 77843, USA
    email: \email{wchiu@cvm.tamu.edu}  
title:
  formatted: "\\pkg{pksensi}: R package to apply global sensitivity analysis in physiologically based kinetic models"
  # If you use tex in the formatted title, also supply version without
  plain:     "pksensi: R package to apply global sensitivity analysis in physiologically based kinetic models"
  # For running headers, if needed
  short:     "**pksensi**: Global SA for PK models in R"
abstract: >
  Sensitivity analysis (SA) is an essential tool for modelers to understand the influence of model parameters on model outputs. It is also increasingly used in developing and assessing pharmacokinetic models. In our earlier work, we applied a global SA workflow to reduce the computational burden in the Bayesian Markov Chain Monte Carlo-based calibration process of a physiologically based pharmacokinetic (PBPK) model. Although several SA algorithms are available, no comprehensive software package exists that allows users to seamlessly solve the PBPK model differential equations, run sensitivity analyses, visualize SA results, and discriminate between the “non-influential” model parameters that can be fixed and those that need calibration. Therefore, we developed an open-source \proglang{R} package, called **pksensi**, to fill this need and make SA more accessible in pharmacological and toxicological research. This package can investigate both parameter uncertainty and sensitivity in pharmacokinetic models, such as PBPK model with multivariate model outputs (multiple time-points and tissue compartments). It also refined the extended Fourier Amplitude Sensitivity Test method for SA by adding a random phase-shift to analyze the statistical variability of the sensitivity index for each model parameter. Furthermore, pksensi includes functions to check the convergence and sensitivity of model parameters, providing a means to assess the robustness of the sensitivity measurement. Utilizing pksensi, we successfully reproduced our previously published SA results of human PBPK modeling for acetaminophen and its two primary metabolites. Overall, **pksensi** improves the user experience of performing SA and can create robust and reproducible results for decision making in pharmacokinetic model calibration. 
keywords:
  # at least one keyword must be supplied
  formatted: [Pharmacokinetics, Physiologically based kinetics, Global sensitivity analysis]
  plain:     [Pharmacokinetics, Physiologically based kinetics, Global sensitivity analysis]
preamble: >
  \usepackage{amsmath}
  \usepackage{booktabs}
bibliography: references.bib 
output: rticles::jss_article

@@

Sensitivity analysis (SA) is a mathematical technique to investigate how variations in model parameters affect model outputs. An increasing number of studies use SA approach to determine which model parameters contribute to high variation in model predictions [@ferretti2016trends]. This technique has also been applied in pharmacology and toxicology research [@loizou2015application; @mcnally2012reconstruction]. Physiologically based kinetic (PK) modeling describes the changes in the concentrations or amounts of a substance within model compartments over time. The goal of SA in PK research is to examine the sensitivity of output variables (e.g. chemical concentration in blood or tissues) in model responds to input parameters, such as anatomical, physiological, and kinetic constants [@mcnally2011workflow]. In addition, this SA approach can be guidance of experimental design, parameter estimation during the model establishment, and qualification processes [@zhang2015sobol]. Through SA, it can optimal experimental design and distinguish the "unidentifiable" parameters that can lead to problems with numerical procedures used in parameter estimation [@chu2010quantitative]. It can be further applied to parameter prioritization and parameter fixing before model calibration [@fphar201800588].

In our earlier work [@fphar201800588], we developed a workflow to apply global SA to reduce the computational burden in the Bayesian, Markov Chain Monte Carlo (MCMC)-based calibration process of a physiologically based pharmacokinetic (PBPK) model. We used \pkg{GNU MCSim} [@JSSv002i09], an effective simulation package for Bayesian population PBPK modeling, to calibrate the model. We found that the extended Fourier Amplitude Sensitivity Test (eFAST), a type of variance-based global SA algorithm, had the best balance of efficiency and accuracy for a complex, multi-compartment, multi-dataset, and multi-metabolite PBPK model. This method was first introduced in @cukier1973study (the classic FAST method with first-order effect only) that study of the sensitivity of coupled reaction systems that are constructed by differential equations with the numerous rate coefficients and then evolved to Sobol' sensitivity measure in @saltelli1999quantitative (the eFAST method with first- and total-order effects). The FAST algorithm creates the multidimensional space of input parameters through "search-curve". It's more computationally efficient in calculating the influence of model parameter than other global SA methods such as the Monte Carlo based approaches from @jansen1999analysis and @owen2013better. Also, we found some efficient visualization approaches that can be used to distinguish between “influential” and “non-influential” parameters. This approach can solve identifiability issue of PBPK model parameter that increases the computational burden of Bayesian analysis, and therefore reduce the computational efficiency. We also found a useful features for communicating the parameter sensitivity in decision making.

@@

Although these tools provide various approaches to perform SA, we have not found a suitable package that provides functions with which users can visualize and distinguish the “non-influential” model parameters and can further apply to parameter fixing in pharmacokinetic and PK modeling. We present here an \proglang{R} package, called \pkg{pksensi}, which is designed to make SA more accessible and reproducible in pharmacological and toxicological researches. This package can investigate both parameter uncertainty and sensitivity in pharmacokinetic and PK model with multivariate model outputs. The types of the kinetic model include the terms "pharmacokinetic", "toxicokinetic", and "biokinetic" model that are exchangeable in the model description. The analytical framework in \pkg{pkseni} not only include global SA but the uncertainty analysis and diagnostic tool to support the modeling process. The design concepts of \pkg{pksensi} are:

@@

We adopted eFAST, a widely used analysis of variance (ANOVA)-like global SA approach to decompose the variance to partial variances for multidimentional parameters in the numerical models, especially for ODE-based dynamical models, in this package. Most of the available eFAST functions, such as the algorithm to generate the parameter space and to set the sampling frequency, were sourced from the \proglang{R} \pkg{sensitivity} package [@R-sensitivity]. The eFAST method computes the sensitivity index (SI) via ANOVA-like decomposition of the function for analysis. This type of global SA approach and SI also know as Sobol' method and Sobol' index. Since the estimated SI is not stable (high variation) when using the lower number of sample size, we included a random phase-shift approach to conduct replicating sampling from random starting points across parameter space to test the convergence and robustness of the sensitivity measurement. Considering the model that is built under $y=f(x_{i})$. The sampling scheme can be written as:

@@

where $x_i$ is the nominal value of the $i$-th parameter, and $\omega_i$ is a vector giving the set of frequencies, one frequency for each parameter, and $\varphi$ is a random phasei-shift coefficient ranged from 0 to $2\pi$. The default set of frequencies is based on the study from @saltelli1999quantitative. Through the random phase-shift, the robust result of the sensitivity measurement should be similar across replications under the same sample size.

To investigate the convergence of SI, we adopted the "exam" approach proposed by [@sarrazin2016global]. This method quantitatively assesses convergence by computing the width of 95% confidence intervals of computed SI for all parameters across all time-points and output variables. Through random phase shift procedure, the SI can repeatly generate with low and high variation based on the given sampling size. Therefore, they can be used to investigate the convergence of SI as

@@

The R package \pkg{pksensi} provides a useful function to solve ODE in pharmacokinetic and PK models using analytical and numerical approaches. The solution can be performed under the pure \proglang{R} programming environment by linking pksensi with the \pkg{deSolve} package [@JSSv033i09]. However, the most efficient way to achieve high computational speed is through compiled, lower-level languages, such as \proglang{FORTRAN}, \proglang{C}, or \proglang{C++}. The \pkg{pksensi} includes a function to compile and create dynamic-link libraries (.dll) on Windows and shared objects (.so) on Unix-liked systems (e.g., Linux and MacOS). This compiled program can load and execute through the build-in function in \proglang{R}. However, Windows users need to install Rtools to compile the source code by using the GNU GCC compiler. More requirements are detailed in @soetaertr. The \proglang{C} implementation of the example model can be found as an example within \pkg{GNU MCSim} [@JSSv002i09]. The \pkg{pksensi} can also link with \pkg{GNU MCSim} to create the model program, used in solving each system of equations.

The uncertainty analysis is a crucial modeling process with SA. Sometimes, it's difficult to have informative data to set up the parameter distribution. The wider range of parameter distribution might cause the computing error at the extreme value. The narrower range might cause the incorrect calibrated result of the model parameter. The uncertainty analysis with Monte Carlo method is an appropriate approach in the data-driven modeling process to prevent this problem. Before fitting the data to model, the Monte Carlo can be applied with the given distribution of model parameter to exam whether the experiment data can be included in the distribution of model output. \pkg{pksensi} also has a function that can link to \pkg{GNU MCSim} to conduct Monte Carlo analysis which is built in \pkg{GNU MCSim} and exam the simulation result. The general workflow of uncertainty and sensitivity analysis is illustrated in Figure \ref{fig:Workflow}.

@@

The output of SA returns a list that contains the given time-points and values of all state variables (e.g., chemical concentrations in blood). This format is particularly suited for further graphical routines within \pkg{pksensi}. Also, time-dependent sensitivity measurements of first and total order effects for all parameters can be plotted simultaneously. The computed SI might show different estimation across replications under the same sample size. Therefore, the inadequate sample number can create an unreliable result and bias. It's essential to use the sufficient sample size to prevent incorrect judgment in parameter fixing. \pkg{pksensi} includes functions to check the convergence and sensitivity of model parameters, providing a means to assess the robustness of the sensitivity measurement. We also developed a "cut-off"-based approach in this package to accurately distinguish between "influential" and "non-influential" parameters. Finally, \pkg{pksensi} provides visualization tools for the effective investigation and communication of results in decision support.

@@

We used the corresponding parameter value of  (APAP) in this example. The parameter value can be generated from \code{parameterize_1comp} function in \pkg{httk} package, which includes physico-chemical properties for over 500 chemicals. The given value of \code{vdist}, \code{ke}, and \code{kgutabs} in \pkg{httk} are 1.1 (L/kg BW), 0.23 (/h), and 2.18 (/h), respectively. The body weight is assumed to be 70 (kg).

```{r, eval=T}
parms <- initParms()
parms["vdist"] <- 1.1
parms["ke"] <- 0.23
parms["kgutabs"] <- 2.18
parms["BW"] <- 70
initState <- initStates(parms=parms)
initState["Agutlument"] <- 10
```

@@

Based on the sensitivity measurement of the total order, the result shows that \code{BW} has a relative lower measurement of SI. However, all parameters do not converge to the setting cut-off, which means the larger sample size is required in further sensitivity testing. Same as \code{plot} function that can assign specific output variable in the examination, the \code{check} function can also use the assignment (\code{vars}) to exam the given output.

@@

In addtion to use \pkg{deSolve} to solve differential equations in PK model, the \pkg{GNU MCSim} can provide better computational efficiency. To solve ODE through \pkg{GNU MCSim}, we need to change the argument to \code{application = mcsim} in \code{compile_model}. The computing time of using \code{solve_fun} in SA is estimated as,

@@

After solving the equations under the same given condition, we can find that \pkg{GNU MCSim} has about 4 - 5 times faster in computing performance than using \pkg{deSolve}. In this case, we only focus on performing the global SA alone for generic PK model without additonal comparison with experiment data. The next example will display and reproduce our previous published result [@fphar201800588] with full global SA workflow.

@@

Same as the example of one-compartment PK model. The model parameter and the corresponding sampling range should be defined to create the parameter matrix. Previously, the probability distributions of model parameters were set to either truncated normal or uniform distribution when the parameters have informative prior information or not. To rapidly reach the acceptance convergence, we apply uniform distribution for all testing parameters. The ranges of informative parameters are set to 1.96-times difference for single side under log-scaled (approximate 54.6 times difference between minimum and maximum in natural scaled). The nominal values of informative model parameters were defined as:

@@

Generally, wide range of parameter value might cause the computing error when solving the differentiial equation. One of the effective ways to prevent this problem is to adjust the value of relative and absolute error tolerance to control the error appearance by resetting these parameters in a lower value. The \code{generate_infile} and \code{solve_mcsim} provide the arguments of \code{rtol} and \code{atol} that adjust the error tolerance to prevent the unwanted error. However, the modification will decrease the computing speed. Therefore, the alternative method to prevent this issue is to detect the crucial parameter range that causes the problem. Also, setting the maximum number of steps to higher value instead of using the default value (500) in \pkg{GNU MCSim} can prevent this problem (internally defined). The maximum number of step is set to 5000 in this case. Here we seperate the global SA of APAP-PBPK model process to several steps. The value can reset through reinstall the the \pkg{GNU MCSim} by using \code{mcsim_install(mxstp = 5000)}.

@@

The 21 testing model parameters are defined in this part, including parameter name, probability distribution, and distributed parameter value. To prevent the computing error, the range of \code{SULT_VmaxC} and \code{UGT_VmaxC} need to adjust from $U(0, 15)$ [@s13318-015-0253-x] to $U(0, 10)$ [@fphar201800588]. The objects \code{q} and \code{dist} are set to the type of distribution that will use to generate the parameter matrix in \pkg{GNU MCSim} (for uncertainty analysis) and \proglang{R} (for SA).

@@

To optimize the computing speed, this case only uses \pkg{GNU MCSim} to estimate the concentration of APAP and its metabolites glucuronide (APAP-G) and sulfate (APAP-S) in plasma. The setting oral dose of APAP is 20 mg/kg in this example. Generally, the input dosing method can be defined through the \code{condition} argument. Since the unit of the given dose is mg/kg, the \code{mgkg_flag} is set to 1. More definition of input schedule functions can be found in the section of input functions in \pkg{GNU MCSim} User’s Manual (https://www.gnu.org/software/mcsim/mcsim.html#Input-functions).

@@

In the setting condition of simulation, The relative and absolute error tolerance (\code{rtol} & \code{atol}) were set to 1e-7 and 1e-9, respectively, to prevent the computing error. The Monte Carlo simulation is run for 1000 iteration as the assignment of \code{monte_carlo}. The input file ('sim.in') and output file ('simmc.out') will be generated under the standard ASCII format.

@@

In addition, through using the \code{check}, the parameter with sensitivity and convergence indices over the given condition can be preliminary detected for all output variables. Based on our previous study, we proposed the heatmap visualization approach \code{heat_check} to distinguish "influential" and "non-influential" parameters with a "cut-off" point. Through the given argument \code{order}, we can select the specific order of sensitivity measurement that we're interested in (Figure \ref{fig:Heat-SI}).

@@

Computational modeling has become an important technique in modern scientific research to support the hypothesis, experiment, and discoveries. Uncertainty and SA is essential knowledge for modelers to conduct the model development and examination. More computational tools had been developed for various types of modeling and simulation. The \pkg{GNU MCSim} is a powerful simulation software package, that can conduct dynamical simulation for biological-based modeling, and it's mainly built for PK research with probabilistic approaches [@bois2009gnu]. The \proglang{R} is a well-developed statistical computing language that can conduct probabilistic programming and have tons of packages for statistical analysis and data visualization. We initiated \pkg{pksensi} that aim to conduct comprehensive and efficient global SA workflow in PK modeling, which hasn't been developed for open source communities. The current version had included \pkg{GNU MCSim} and \pkg{deSolve} as the main solver in our model assessment. Compare with \pkg{deSolve} package, we found that The \pkg{GNU MCSim} can provide better efficiency to solve the ODE in PK models. Although \pkg{pksensi} provide the essential features and functions to install and link with \pkg{GNU MCSim}, conducting the uncertainty and SA in PK modeling, the overall learning curve of this workflow might steep for users that are not familiar with \pkg{GNU MCSim} from model development, debugging, and testing. Therefore, the general \proglang{R} package, such as \pkg{deSolve}, can be an alternative option for \proglang{R} user. Some developed package in \proglang{R} had also provide the features to solve ODE such as \pkg{RxODE} [@wang2016tutorial] and \pkg{mrgsolve} [@R-mrgsolve]. We hope to incorporate these packages that can improve the user experience in using \pkg{pksensi} in the future.

Moreover, some fundamental packages on the CRAN repository and other software platforms had been built to perform global SA with different algorithms. The development of \pkg{pksensi} is mainly for PK models that are constructed by ODE. It provides a straightforward application to investigate the impact of parameters on model output across simulation time-points and variables. We adopted the eFSAT method, a common variance-based SA that had better efficiency and accuracy in the model assessment. It can exam the parameter effect from first order to the total order. In addition, the package can not only conduct global SA but also include the function to assess the convergence of sensitivity measurements, which is rarely discussed in most global SA studies and its software package. We also develop the function to visualize the output result that can help us distinguish the "influential" and "non-influential" parameters that can be applied in parameter fixing in the model Calibration. In addition to eFAST, Some studies applied another global SA algorithm in the physiologically and biologically based ODE models, such as compartmental absorption and transit (also known as CAT) and other system biology models [@melillo2019variance; @sumner2012methodology]. Base on our knowledge, the eFAST is a least computational expensive approch than other Monte-Carlo based Sobol' method. In future version, we expect to add additional global SA approach that had been applied in advanced studies [@glen2012estimating; @ring2017identifying]. The other features such as principle component analysis [@lamboni2011multivariate] were considered to further include in our global SA workflow.

It is worth noting that the global SA results are limited to the specific model and the given range of the model parameter. The provided conditions, such as the number of examing parameter, parameter distribution, parameter range, and the given dosing level in PK model might have a potential impact on the level of importance of each parameter. The misconduct of the setting condition might lead to the incorrect estimated result. In other words, if we have various types of experiment data (e.g. different exposure routes and biological measurements) and will use them in model calibration, it will enhance the difficulties in global SA. The complex model with a high dimension of the parameter is also a challenge in global SA, which will need higher computer demand in the model assessment. More additional applications still need to be investigated and strengthen. Overall, \pkg{pksensi} provide an additional option of performing global SA for PK model and can create robust and reproducible results for decision making in model calibration. Although \pkg{pksensi} used here is mainly for PK model, it can also work for the other ODE-based dynamic model to investigate the sensitivity of model output from the input parameters.

>>>>

---
author:
  - name: Nan-Hung Hsieh
    affiliation: Texas A&M University
  - name: Brad Reisfeld
    affiliation: Colorado State University 
  - name: Weihsueh A. Chiu
    affiliation: Texas A&M University
    address: >
      4458 TAMU, 
      College Station, TX 77843, USA
    email: \email{wchiu@cvm.tamu.edu}  
title:
  formatted: "\\pkg{pksensi}: R package to apply global sensitivity analysis in physiologically based kinetic models"
  # If you use tex in the formatted title, also supply version without
  plain:     "pksensi: R package to apply global sensitivity analysis in physiologically based kinetic models"
  # For running headers, if needed
  short:     "**pksensi**: Global SA for PK models in R"
abstract: >
  Sensitivity analysis (SA) is an essential tool for modelers to understand the influence of model parameters on model outputs. It is also increasingly used in developing and assessing pharmacokinetic models. For instance, we recently applied a global SA workflow to reduce the computational burden in the Bayesian Markov Chain Monte Carlo-based calibration process of a physiologically based pharmacokinetic (PBPK) model. Although several SA algorithms and software packages are available, no comprehensive software package exists that allows users to seamlessly solve the PBPK model differential equations, run sensitivity analyses, visualize SA results, and discriminate between the “non-influential” model parameters that can be fixed and those that need calibration. Therefore, we developed an open-source \proglang{R} package, called **pksensi**, to fill this need and make SA more accessible in pharmacological and toxicological research. This package can investigate both parameter uncertainty and sensitivity in pharmacokinetic models, such as PBPK model with multivariate model outputs (multiple time-points and tissue compartments). It also refined the extended Fourier Amplitude Sensitivity Test method for SA by adding a random phase-shift to analyze the statistical variability of the sensitivity index for each model parameter. Furthermore, pksensi includes functions to check the convergence and sensitivity of model parameters, providing a means to assess the robustness of the SA results. Utilizing pksensi, we successfully reproduced our previously published SA results of human PBPK modeling for acetaminophen and its two primary metabolites. Overall, **pksensi** improves the user experience of performing SA and can create robust and reproducible results for decision making in pharmacokinetic model calibration. 
keywords:
  # at least one keyword must be supplied
  formatted: [Pharmacokinetics, Physiologically based kinetics, Global sensitivity analysis]
  plain:     [Pharmacokinetics, Physiologically based kinetics, Global sensitivity analysis]
preamble: >
  \usepackage{amsmath}
  \usepackage{booktabs}
bibliography: references.bib 
output: rticles::jss_article

@@

Sensitivity analysis (SA) is a mathematical technique to investigate how variations in model parameters affect model outputs [@saltelli2008global]. An increasing number of studies use SA approach to determine which model parameters contribute to high variation in model predictions [@ferretti2016trends]. This technique has also been applied in pharmacology and toxicology research [@loizou2015application; @mcnally2012reconstruction]. Pharmacokinetic models describe the changes in the concentrations or amounts of a substance in the body over time (the various terms for the kinetic models, including "pharmacokinetic", "toxicokinetic", and "biokinetic" models, are used exchangeably here). The goal of SA in pharmacokinetic research is to examine the sensitivity of output variables (e.g. chemical concentration in blood or tissues) in model responds to input parameters, such as anatomical, physiological, and kinetic constants [@mcnally2011workflow]. In addition, this SA approach can be guidance of experimental design, parameter estimation during the model establishment, and qualification processes [@zhang2015sobol]. Through SA, it can optimal experimental design and distinguish the "unidentifiable" parameters that can lead to problems with numerical procedures used in parameter estimation [@chu2010quantitative]. It can be further applied to parameter prioritization and parameter fixing before model calibration [@fphar201800588].

In our earlier work [@fphar201800588], we developed an updated workflow to apply global SA to reduce the computational burden in the Bayesian, Markov Chain Monte Carlo (MCMC)-based calibration process of a physiologically based pharmacokinetic (PBPK) model. We used \pkg{GNU MCSim} [@JSSv002i09], an effective simulation package for Bayesian population PBPK modeling, to calibrate the model. We found that the extended Fourier Amplitude Sensitivity Test (eFAST), a type of variance-based global SA algorithm, had the best balance of efficiency and accuracy for a complex, multi-compartment, multi-dataset, and multi-metabolite PBPK model. This method was first introduced in @cukier1973study (the classic FAST method with first-order effect only) that study of the sensitivity of coupled reaction systems that are constructed by differential equations with the numerous rate coefficients and then evolved to Sobol' sensitivity measure in @saltelli1999quantitative (the eFAST method with first- and total-order effects). The FAST algorithm creates the multidimensional space of input parameters through "search-curve". It's more computationally efficient in calculating the influence of model parameter than other global SA methods such as the Monte Carlo based approaches from @jansen1999analysis and @owen2013better. Our previous work also found some efficient visualization approaches that can be used to distinguish between “influential” and “non-influential” parameters. This approach address common identifiability issues of PBPK model parameter that increase the computational burden of Bayesian analysis [@garcia2015identifiability], and therefore reduce the computational efficiency. We developed approaches for communicating parameter sensitivity in decision making.

@@

Although these tools provide various approaches to perform SA, they each individually have limitations for application to pharmacokinetic and PK modeling. We present here an \proglang{R} package, called \pkg{pksensi}, which is designed to make SA more accessible and reproducible in pharmacological and toxicological researches. This package can investigate both parameter uncertainty and sensitivity in pharmacokinetic and PK model with multivariate model outputs. Moreover, it seamlessly integrates with both general ODE solvers available in R or as external \proglang{C} code, as well as the \pkg{GNU MCSim} software used for conducting sophisticated PK modeling, including MCMC. The analytical framework in \pkg{pkseni} not only include global SA but the uncertainty analysis and diagnostic tool to support the modeling process. The design concepts of \pkg{pksensi} are:

@@

We adopted eFAST, a widely used analysis of variance (ANOVA)-like global SA approach to decompose the variance to partial variances for multidimentional parameters in the numerical models, especially for ODE-based dynamical models, in this package. Most of the available eFAST functions, such as the algorithm to generate the parameter space and to set the sampling frequency, were sourced from the \proglang{R} \pkg{sensitivity} package [@R-sensitivity]. The eFAST method computes the sensitivity index (SI) via ANOVA-like decomposition of the function for analysis. This type of global SA approach and SI also know as Sobol' method and Sobol' index. Since the estimated SI is not stable (high variation) when using smaller sample sizes, we included a random phase-shift approach to conduct replicating sampling from random starting points across parameter space to test the convergence and robustness of the sensitivity measurement. Considering the model that is built under $y=f(x_{i})$. The sampling scheme can be written as:

@@

where $x_i$ is the nominal value of the $i$-th parameter, and $\omega_i$ is a vector giving the set of frequencies, one frequency for each parameter, and $\varphi$ is a random phasei-shift coefficient ranged from 0 to $2\pi$. The default set of frequencies is based on the study from @saltelli1999quantitative. Through the random phase-shift, the robust result of the sensitivity measurement should be similar across replications under the same sample size.
For most Sobol'-based SA algorithm, the larger sample size is needed to reduce the probability of having negative estimates. However, FAST estimates are always positive according to its construction.

To investigate the convergence of SI, we adopted the "exam" approach proposed by [@sarrazin2016global]. This method quantitatively assesses convergence by computing the width of 95% confidence intervals of computed SI for all parameters across all time-points and output variables. Through random phase shift procedure, the replicate SI values can be repeatedly sampled independently. Therefore, they can be used to investigate the convergence of SI as

@@

The R package \pkg{pksensi} provides a useful function to solve ODEs in pharmacokinetic and PK models using analytical and numerical approaches. The solution can be performed under the pure \proglang{R} programming environment by linking pksensi with the \pkg{deSolve} package [@JSSv033i09]. However, the most efficient way to achieve high computational speed is through compiled, lower-level languages, such as \proglang{FORTRAN}, \proglang{C}, or \proglang{C++}. The \pkg{pksensi} includes a function to compile and create dynamic-link libraries (.dll) on Windows and shared objects (.so) on Unix-liked systems (e.g., Linux and MacOS). This compiled program can load and execute through the build-in function in \proglang{R}. However, Windows users need to install Rtools to compile the source code by using the GNU GCC compiler. More requirements are detailed in @soetaertr. The \proglang{C} implementation of the example model can be found as an example within \pkg{GNU MCSim} [@JSSv002i09]. The \pkg{pksensi} can also link with \pkg{GNU MCSim} to create the model program, used in solving each system of equations.

The uncertainty analysis is a crucial modeling process within SA. Sometimes, it's difficult to have informative data to set up the parameter distribution. A wider range for a parameter distribution might cause a numerical error at extreme values. On the other hand, narrower ranges might cause the incorrect calibrated result of the model parameter. The uncertainty analysis using Monte Carlo is an appropriate approach in the data-driven modeling process to address this problem. Before fitting the data to model, the Monte Carlo can be applied with the given distribution of model parameter to examine whether the experiment data within the range of in the distribution of model output. \pkg{pksensi} also has a function that can link to \pkg{GNU MCSim} to conduct Monte Carlo analysis which is built in \pkg{GNU MCSim} and exam the simulation result. The general workflow of uncertainty and sensitivity analysis is illustrated in Figure \ref{fig:Workflow}.

@@

The output of SA returns a list that contains the given time-points and values of all state variables (e.g., chemical concentrations in blood). This format is particularly suited for further graphical routines within \pkg{pksensi}. Also, time-dependent sensitivity measurements of first and total order effects for all parameters can be plotted simultaneously. The computed SI might show different estimation across replications under the same sample size. Therefore, the inadequate sample number can create an unreliable result and bias. It is essential to use a sufficient sample size to prevent incorrect judgment in parameter fixing. \pkg{pksensi} includes functions to check the convergence and sensitivity of model parameters, providing a means to assess the robustness of the sensitivity measurement. We also developed a "cut-off"-based approach in this package to accurately distinguish between "influential" and "non-influential" parameters. Finally, \pkg{pksensi} provides visualization tools for the effective investigation and communication of results in decision support.

@@

We used the corresponding parameter value of acetaminophen (APAP) in this example. These model parameters are derived from the *in-vivo* or *in-vitro* experiment results. The parameter value can be generated from \code{parameterize_1comp} function in \pkg{httk} package, which includes physico-chemical properties for over 500 chemicals. The given value of \code{vdist}, \code{ke}, and \code{kgutabs} in \pkg{httk} are 1.1 (L/kg BW), 0.23 (/h), and 2.18 (/h), respectively. The body weight is assumed to be 70 (kg).
```{r, eval=T}
parms <- initParms()
parms["vdist"] <- 1.1
parms["ke"] <- 0.23
parms["kgutabs"] <- 2.18
parms["BW"] <- 70
initState <- initStates(parms=parms)
initState["Agutlument"] <- 10
```

@@

Based on the sensitivity measurement of the total order, the result shows that \code{BW} has a relative lower measurement of SI. However, all parameters do not converge to the setting cut-off, which means the larger sample size is required in further sensitivity testing. Similiar to the \code{plot} function that can assign specific output variable in the examination, the \code{check} function can also use the assignment (\code{vars}) to exam the given output.

@@

In addition to using \pkg{deSolve} to solve differential equations in PK model, the \pkg{GNU MCSim} can also be used, and provides better computational efficiency. To solve ODEs through \pkg{GNU MCSim}, we need to change the argument to \code{application = mcsim} in \code{compile_model}. The computing time of using \code{solve_fun} in SA is estimated as,

@@

After solving the equations under the same given condition, we can find that \pkg{GNU MCSim} has about 4 - 5 times faster in computing performance than using \pkg{deSolve}. In this case, we only focus on performing the global SA alone for generic PK model without additional comparison with experimental data. The next example will display and reproduce our previous published result [@fphar201800588] with full global SA workflow.

@@

As in the example of one-compartment PK model, the model parameter and the corresponding sampling range should be defined to create the parameter matrix. Previously, the probability distributions of model parameters were set to either truncated normal or uniform distributions when the parameters have informative prior information or not, repectively. To rapidly reach the acceptance convergence in  this example, we apply uniform distribution for all testing parameters. The ranges of informative parameters are set to 1.96-times difference for single side under log-scaled (approximate 54.6 times difference between minimum and maximum in natural scaled). The nominal values of informative model parameters were defined as:

@@

Generally, wide range of parameter value might cause the computing error when solving the differential equation. One of the effective ways to prevent this problem is to adjust the value of relative and absolute error tolerance to control the error appearance by resetting these parameters in a lower value. The \code{generate_infile} and \code{solve_mcsim} provide the arguments of \code{rtol} and \code{atol} that adjust the error tolerance to prevent the unwanted error. However, the modification will decrease the computing speed. Therefore, the alternative method to prevent this issue is to detect the crucial parameter range that causes the problem. Also, setting the maximum number of steps to higher value instead of using the default value (500) in \pkg{GNU MCSim} can prevent this problem (internally defined). The maximum number of step is set to 5000 in this case. Here we separate the global SA of APAP-PBPK model process to several steps. The value can reset through reinstall the the \pkg{GNU MCSim} by using \code{mcsim_install(mxstp = 5000)}.

@@

The 21 tested model parameters are defined in this step, including parameter name, probability distribution, and distributed parameter value. To prevent errors, the range of \code{SULT_VmaxC} and \code{UGT_VmaxC} needed to be adjusted from $U(0, 15)$ [@s13318-015-0253-x] to $U(0, 10)$ [@fphar201800588]. The objects \code{q} and \code{dist} are set to the type of distribution that will use to generate the parameter matrix in \pkg{GNU MCSim} (for uncertainty analysis) and \proglang{R} (for SA).

@@

To optimize the computing speed, this case only uses \pkg{GNU MCSim} to estimate the concentration of APAP and its metabolites glucuronide (APAP-G) and sulfate (APAP-S) in plasma. The oral dose of APAP is 20 mg/kg in this example. Generally, the input dosing method can be defined through the \code{condition} argument. Since the unit of the given dose is mg/kg, the \code{mgkg_flag} is set to 1. More definition of input schedule functions can be found in the section of input functions in \pkg{GNU MCSim} User’s Manual (https://www.gnu.org/software/mcsim/mcsim.html#Input-functions).

@@

In the setting condition of simulation, The relative and absolute error tolerance (\code{rtol} & \code{atol}) were set to 1e-7 and 1e-9, respectively, to prevent the numerical errors. The Monte Carlo simulation is run for 1000 iterations in the assignment of \code{monte_carlo}. The input file ('sim.in') and output file ('simmc.out') will be generated under the standard ASCII format.

@@

In addition, through using \code{check}, the parameter with sensitivity and convergence indices over the given condition can be preliminary detected for all output variables. Based on our previous study, we proposed the heatmap visualization approach \code{heat_check} to distinguish "influential" and "non-influential" parameters with a "cut-off" point. Through the given argument \code{order}, we can select the specific order of sensitivity measurement that we're interested in (Figure \ref{fig:Heat-SI}).

@@

Computational modeling has become an important technique in modern scientific research to support hypotheses, experiments, and discoveries.  Uncertainty and SA are essential tools for modelers in the conduct of model development and examination. Several packages on the CRAN repository and other software platforms had been built to perform global SA with a variety of algorithms, but none have included integration with ODE-based models, such as PK models.

We initiated \pkg{pksensi} that aim to conduct comprehensive and efficient global SA workflow in PK modeling, filling a key gap for open source modeling communities. It provides a straightforward application to investigate the impact of parameters on model outputs across simulation time-points and variables. We adopted the eFSAT method, a common variance-based SA that had better efficiency and accuracy in the model assessment. It can exam the parameter effect from first order to the total order. In addition, the package can not only conduct global SA but also include the function to assess the convergence of sensitivity measurements, which is rarely discussed in most global SA studies and its software package. We also develop the function to visualize the output result that can help us distinguish the "influential" and "non-influential" parameters that can be applied in parameter fixing in the model calibration. 
In addition to eFAST, Some studies applied another global SA algorithm in the physiologically and biologically based ODE models, such as compartmental absorption and transit (also known as CAT) and other system biology models [@melillo2019variance; @sumner2012methodology].  Base on our experience, eFAST is the least computationally expensive approach compared to other Monte-Carlo based Sobol’ method.  In the future version, we expect to add additional global SA approach that had been applied in advanced studies [@glen2012estimating; @ring2017identifying]. The other features such as principle component analysis [@lamboni2011multivariate] were considered to further include in our global SA workflow.

We chose to integrate sensitivity analysis with \pkg{GNU MCSim} because it is a powerful simulation software package for dynamical simulation of biological-based models, mainly built for PK research with probabilistic approaches [@bois2009gnu] and had developed over decades. In addition, the current version of \pkg{pksensi} includes both \pkg{GNU MCSim} and \pkg{deSolve} as the main ODE solvers.  Compared with the \pkg{deSolve} package, the \pkg{GNU MCSim} can provide better speed and efficiency to solve the ODEs in PK models. Although \pkg{pksensi} provides the essential features and functions to install and link with \pkg{GNU MCSim}, conducting the uncertainty and SA in PK modeling, the overall learning curve of this workflow might steep for users that are not familiar with \pkg{GNU MCSim} from model development, debugging, and testing. Therefore, the general \proglang{R} package, such as \pkg{deSolve}, can be an alternative option for R users. Some other packages in R are also available to solve ODEs, such as \pkg{RxODE} [@wang2016tutorial] and \pkg{mrgsolve} [@R-mrgsolve]. We hope to incorporate these packages to expand the flexibility of using \pkg{pksensi} in the future.

It is worth noting that the global SA results are limited to the specific model and the given range of the model parameter. The provided conditions, such as the number of parameters examined, the parameter distributions/ranges, and the given dosing level in PK model might have a potential impact on the level of importance of each parameter. In other words, if we have various types of experiment data (e.g. different exposure routes and biological measurements) and will use them in model calibration, it will enhance the difficulties in global SA. The complex model with a high dimension of the parameter is also a challenge in global SA, which will need higher computer demand in the model assessment. As approaches are developed to address these complexities, we hope to incorporate them into \pkg{pksensi}.

Overall, \pkg{pksensi} provides comprehensive features and functions for performing global SA for PK models and can create robust and reproducible results for decision making in model calibration. Although \pkg{pksensi} used here is mainly for pharmacokinetic modeling, it can also work for the other ODE-based dynamic models to investigate the sensitivity of model output from the input parameters.
